Waking up a sleeping rabbit:\\ On Natural-Language Sentence Generation with FF

### 2009-12-19 trunk/Papers/crisp-ff-10/conclusion.tex
4 pages. whew.

Generation of natural language sentences is interesting, as a planning
problem, both in its application value and in the challenges it
poses. Our observations herein have above all the message that, as of
now, FF is a practical option for this application. It is baffling
with which ease this milestone came about, and certainly this suggests
to look more closely at automatic planner configuration techniques.

>>>>

Generation of natural language sentences is interesting as a planning
problem, both in its application value and in the challenges it
poses. After the changes described in this paper, FF is a practical
option for this application. It is striking how simple these changes
were, and certainly this suggests to look more closely at automatic
planner configuration techniques.

### 2009-12-19 trunk/Papers/crisp-ff-10/crisp-ff.tex
4 pages. whew.

%It turns out that, as far as the suitability of commonly used
%pre-processing and search techniques are concerned, sentence
%generation diverges significantly from most known benchmark
%domains. We examine this here, in detail, for the case of FF
%\cite{HoffmannNebel01}. We show how certain hitherto irrelevant
%details of the pre-processing suddenly become deadly obstacles to
%performance, and how certain widely used search techniques -- namely
%goal orderings and the relaxed plan heuristic -- are completely off
%the mark and partly even detrimental. For some of the observed
%pitfalls, we are able to provide simple fixes; others pose serious
%challenges to planning research. We focus first on the pre-processing,
%then on the search techniques.

It turns out that FF's exhibits, in this domain, a number of
detrimental phenomena that were previously unheard of. We now examine
these in some detail.  For some of the observed pitfalls, we are able
to provide simple fixes. Others pose serious challenges to planning
research.

%We focus first on the pre-processing,
%then on the search techniques.

\subsection{Pre-Processing}
\label{sec:crisp-ff:preprocess}

@@

FF's pre-process contains two details that, for a long time until the
writing of this paper, were so detrimental that we were not able to
seriously propose FF as a practical solution.
%Simple fixes
%to these details are so effective that FF is now a practical (if not
%perfect) possibility.

While pre-processing the ADL formulas, FF has a sub-routine that
checks the grounded formulas
% -- quantifiers already compiled into
%grounded conjunctions and disjunctions -- 
for tautologies. This involves a loop checking all pairs of
sub-formulas within every conjunction and disjunction, testing whether
they are identical/identical modulo their sign. For example, if we
find $\phi$ and $\neg \phi$ within a disjunction, then the disjunction
is replaced by TRUE. Now, we haven't tested whether this sub-routine
has a notable effect on the formulas in other benchmarks. It certainly
has no such effect in our domain, other than taking a huge amount of
runtime on the long goal conjunctions stating that we do not wish to
have any open substitution nodes nor any distractors. The fix is to
switch this sub-routine off. This trivial operation alone can render
previously infeasible instances (pre-processing time $>15$ minutes)
harmless (instance solved in $6$ seconds {\em total} time).

@@

FF's pre-process contains a very small imprudence that apparently
didn't surface in other domains, but that does in ours. FF
distinguishes operators into ``easy'' (DNF precondition) and ``hard''
(non-DNF precondition), which go through different instantiation
procedures. The procedure for hard operators uses a data structure
encompassing an integer for every possible instantiation of all
predicates. Since this does not take into account reachability, it is
prohibitively huge in our domain. Now, we actually have no hard
operators. But FF builds the data structure anyway. This sometimes
results in a couple of seconds runtime overhead; it also sometimes
exhausts memory and causes a segmentation fault.

@@

FF's pre-processing pitfalls -- and the fact that they didn't occur to
anyone before -- are certainly baffling, although easily fixed. We now
turn our attention to some of FF's search techniques, variations of
which are used in many planners and which are very much not easily
fixed.

@@

The goal agenda is completely off-target in our domain. The most
striking point concerns the interaction between the goal to create a
sentence ($\neg \mathsf{subst}(S,\mathsf{root})$ in the rabbit
example), and the goal to express a certain meaning
($\mathsf{sleep}(e,r_1)$ in the rabbit example). The goal agenda
orders the former before the latter because it detects that, once a
meaning is expressed, the sentence creation has started and it is not
possible to start anew anymore. Unless the inverse also
holds,\footnote{The inverse holds if, in the given grammar, the only
  way to create a meaning is to create a full sentence -- i.e., there
  are no side sentences like ``the sleeping rabbit''.}
$\mathsf{subst}(S,\mathsf{root})$ is alone in the first goal agenda
entry, i.e., there is no goal stating the actual meaning the sentence
is supposed to have. Consequently, the planner decides to generate
{\em any} sentence. If it happens to be the wrong sentence, the
planner is stuck in a dead-end. %
%While that dead end is recognized by
%the relaxed plan heuristic -- i.e., it is easy to see that the actions
%needed for achieving the goal cannot be applied anymore -- certainly
%this is an entirely unintended (and a bit amusing) outcome of this
%technique.
% Switching it off, we {\bf \dots
%example before/after.}
It is not clear to us, at this point, how goal ordering techniques
could be modified so that they return more sensible results.

@@

In our improved FF, the goal agenda is simply switched off. Having
disposed of goal orderings in that way, let us turn our attention to
the relaxed plan heuristic. %
%While the goal agenda meltdown may be more amusing than alarming, we
%next report that relaxed plan heuristics are not far from such a
%meltdown either. They are fatally 
This meets its Waterloo as well, due to the weird lifecycle of facts
encoding substitution nodes and distractors. In instances where these
constructs are important for the construction of a valid sentence,
these facts are required to be false in the goal, and most of them are
initially false. However, any valid plan must make many of them true
at some intermediate time point (e.g.\ distractors arise depending on
the decisions taken by the planner). Hence, from the perspective of
the relaxed plan heuristic, from one state to the next whole classes
of goals suddenly appear ``out of the blue''. 
%These goals are not
%accounted for at all in the preceding state.
%, and hence the relaxed plan is
%completely un-informative.

Consider again the rabbit example. In the initial state, the relaxed
plan has length $1$ because we only need to achieve $\neg
\mathsf{subst}(S,\mathsf{root})$ and $\mathsf{sleep}(e,r_1)$, which is
done by the single action ``sleeps''. However, that action has the
effect ``forall (?y - individual) (when (not (= ?y ?xself))
(distractor ?subjnode ?y))'' that introduces new distractors. Those
must be removed, so they form ``new'' goals. In the given case, the
relaxed plan now contains two actions to achieve these goals. In
general (if there are $n$ rabbits, say), the relaxed plan may become
arbitrarily long.
%To give just one concrete example, say we are in the initial state,
%and have not yet committed to a sentence. For certain kinds of
%sentences ({\bf more concrete here???}), the relaxed plan has length
%$1$ because the we only need to choose the desired meaning. However,
%in the state $s$ resulting from this action, the mechanisms of the
%planning task introduce all the sub-goals that are required in order
%to pin-point that meaning (in particular we get a new goal for all the
%distractors that must be removed). Hence the relaxed plan is much
%longer than before -- arbitrarily longer, in general.  
Given FF's particular search algorithm, enforced hill-climbing (EHC),
this means that {\em FF turns into a blind breadth-first search}: EHC
starts at $h$ value $1$ and thereafter needs to solve the whole
problem in order to find a (goal) state with lower $h$ value.

@@

More generally, from this example we immediately see that, in the
terms of Hoffmann \shortcite{hoffmann:jair-05}, the exit distance from
local minima under $h^+$ is unbounded. Also, we have unrecognized
dead-ends, i.e., dead-end states that have a relaxed plan. An example
is the initial state of a task that features distractors for which
there is no means of removal -- ignoring the distractors, the initial
$h$ value will still be $1$. So our domain is in the most difficult
class of Hoffmann's \shortcite{hoffmann:jair-05} ``planning domain
taxonomy''. While it shares this property with several other
challenging domains, the extremity of turning EHC into a breadth-first
search is unheard of.

@@

Can we design heuristics that are better at dealing with the lifecycle
of substitution nodes and distractors? It is easy to see that the
problem of not noticing them in the initial state is present for most
known heuristics
\cite{helmert:icaps-04,richter:etal:aaai-08,karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09,cai:etal:icaps-09}. The
two notable eceptions are the $h^m$ family, and abstraction
heuristics. We ran a few tests with an example like the rabbit one,
except that there are 10 distractors. The Graphplan estimate for the
initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
$4$ is far away from the real goal distance $14$. With merge\&shrink
\cite{helmert:etal:icaps07}, we obtained the estimate $11$ for the
initial state, which is much better. Still the search space with that
latter heuristic remained fairly large ($5200$ nodes vs.\ $13800$
nodes for blind search) and it is not clear how well the technique
will scale to larger instances and instances with a more complex
grammar. Certainly, interesting challenges remain.

@@

The only FF technique that appears to work reasonably well is its
action pruning technique, the ``helpful actions''. While the relaxed
plan is often much too short, it appears to always contain at least
one of the actions that actually are useful. As our experiments data
will show, it certainly prunes the set of available actions quite
effectively.

%For example, in the
%10-distractor test instance, $16875$ nodes are explored without
%helpful actions pruning; with pruning, this reduces to $3078$. For
%larger examples, the factor between the two numbers tends to become
%larger.

%Since EHC is often completely inadequate, and helpful actions are
%quite effective,

Given the above observations regarding EHC and helpful actions, our
modified FF directly runs best-first search with helpful actions (as
opposed to original FF which switches to best-first with{\em out}
helpful actions if EHC fails).

>>>>

FF exhibits a number of problems in this domain that were previously
unknown.  The basic observation of FF's inefficiency on these inputs
was made before
\cite{koller10:_exper_with_plann_for_natur_languag_gener}; here we
examine its causes in detail. For some of the observed pitfalls, we
are able to provide simple fixes. Others pose serious challenges to
planning research.

\subsection{Preprocessing}
\label{sec:crisp-ff:preprocess}

@@

FF's preprocessor contains two details that, before this paper, were
so detrimental that we were not able to seriously propose FF as a
practical solution.
%Simple fixes
%to these details are so effective that FF is now a practical (if not
%perfect) possibility.

While preprocessing the ADL formulas, FF has a subroutine that
checks the grounded formulas
% -- quantifiers already compiled into
%grounded conjunctions and disjunctions -- 
for tautologies. This involves a loop checking all pairs of
sub-formulas within every conjunction and disjunction, testing whether
they are identical/identical modulo their sign. For example, if we
find $\phi$ and $\neg \phi$ within a disjunction, then the disjunction
is replaced by TRUE. We haven't tested whether this subroutine has a
notable effect in other benchmarks. In our domain, the only effect it
has is to take a huge amount of runtime on the long goal conjunctions
stating that we do not want any open substitution nodes or
distractors. The fix is to switch this subroutine off. This trivial
operation alone can render previously infeasible instances
(preprocessing time $>15$ minutes) harmless (instance solved in $6$
seconds {\em total} time).

@@

FF's preprocessor contains a very small imprudence that apparently
didn't surface in other domains, but that does in ours. FF
distinguishes operators into ``easy'' (DNF precondition) and ``hard''
(non-DNF precondition), which go through different instantiation
procedures. The procedure for hard operators uses a data structure
encompassing an integer for every possible instantiation of all
predicates. Since this does not take into account reachability, it is
prohibitively huge in our domain. Now, we actually have no hard
operators. But FF builds the data structure anyway, which exhausts
memory and causes a segmentation fault on serious instances.

@@

FF's preprocessing pitfalls -- and the fact that they didn't occur to
anyone before -- are certainly baffling, although easily fixed. We now
turn our attention to some of FF's search techniques, variations of
which are used in many planners and which are very much not easily
fixed.

@@

The goal agenda is completely off-target in our domain. The most
striking point concerns the interaction between the goal to create a
sentence ($\neg \mathsf{subst}(S,\mathsf{root})$ in the rabbit
example), and the goal to express a certain meaning
($\f{expressed}(\f{sleep},e,r_1)$). The goal agenda orders the former
before the latter because it detects that, once a meaning is
expressed, the sentence creation has started and it is not possible to
start anew anymore. Unless the inverse also holds,\footnote{The
  inverse holds if, in the given grammar, the only way to communicate
  a meaning is in the main clause -- i.e., there are no subordinate
  clauses like ``the sleeping rabbit''.}
$\mathsf{subst}(S,\mathsf{root})$ is alone in the first goal agenda
entry. Consequently, the planner decides to generate {\em any}
sentence, with some arbitrary meaning. If it happens to be the wrong
sentence, the planner is stuck in a dead-end. %
%While that dead end is recognized by
%the relaxed plan heuristic -- i.e., it is easy to see that the actions
%needed for achieving the goal cannot be applied anymore -- certainly
%this is an entirely unintended (and a bit amusing) outcome of this
%technique.
% Switching it off, we {\bf \dots
%example before/after.}
It is not clear to us, at this point, how goal ordering techniques
could be modified so that they return more sensible results.

@@

In our improved FF, the goal agenda is simply switched off. Having
disposed of goal orderings in that way, let us turn our attention to
the relaxed plan heuristic. %
%While the goal agenda meltdown may be more amusing than alarming, we
%next report that relaxed plan heuristics are not far from such a
%meltdown either. They are fatally 
This meets its Waterloo as well, due to the unusual lifecycle of facts
encoding substitution nodes and distractors. 
%In instances where these
%constructs are important for the construction of a valid sentence,
These facts are required to be false in the goal, and most of them are
initially false. However, any valid plan must make many of them true
at some intermediate time point (e.g.\ distractors arise depending on
the decisions taken by the planner). Hence, from the perspective of
the relaxed plan heuristic, from one state to the next whole classes
of goals suddenly appear ``out of the blue''.
%These goals are not
%accounted for at all in the preceding state.
%, and hence the relaxed plan is
%completely un-informative.

Consider again the rabbit example. In the initial state, the relaxed
plan has length $1$ because we only need to achieve $\neg
\mathsf{subst}(S,\mathsf{root})$ and $\f{expressed}(\f{sleep},e,r_1)$,
which is done by the single action ``sleeps''. However, that action
has the effect $\forall y. y \neq x_1 \rightarrow
\f{distractor}(u_1,y)$, which introduces new distractors. Thus the
relaxed plan for the second state contains \emph{two} actions. In
general (if there are $n$ rabbits, say), the relaxed plan may become
arbitrarily long.
%To give just one concrete example, say we are in the initial state,
%and have not yet committed to a sentence. For certain kinds of
%sentences ({\bf more concrete here???}), the relaxed plan has length
%$1$ because the we only need to choose the desired meaning. However,
%in the state $s$ resulting from this action, the mechanisms of the
%planning task introduce all the sub-goals that are required in order
%to pin-point that meaning (in particular we get a new goal for all the
%distractors that must be removed). Hence the relaxed plan is much
%longer than before -- arbitrarily longer, in general.  
For FF's particular search algorithm, enforced hill-climbing (EHC),
this means that {\em FF turns into a blind breadth-first search}: EHC
starts at $h$ value $1$ and thereafter needs to solve the whole
problem in order to find a (goal) state with lower $h$ value.

@@

More generally, from this example we immediately see that, in the
terms of Hoffmann \shortcite{hoffmann:jair-05}, the exit distance from
local minima under $h^+$ is unbounded. Also, we have unrecognized
dead-ends, i.e., dead-end states that have a relaxed plan. An example
is the initial state of a generation problem that it unsolvable
because not all distractors can be removed; its $h$ value will still
be $1$. So our domain is in the most difficult class of Hoffmann's
\shortcite{hoffmann:jair-05} ``planning domain taxonomy''. While it
shares this property with several other challenging domains, the
extremity of turning EHC into a breadth-first search is unheard of.

@@

Can we design heuristics that are better at dealing with the lifecycle
of substitution nodes and distractors? It is easy to see that the
problem of not noticing them in the initial state is present for most
known heuristics
\cite{helmert:icaps-04,richter:etal:aaai-08,karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09,cai:etal:icaps-09}. The
two notable eceptions are the $h^m$ family, and abstraction
heuristics. We ran a few tests with a variant of the example problem
involving 10 rabbits.  The Graphplan estimate for the initial state is
$4$, i.e., $h^2$ is better than $h^+$ here (although still far from
the real goal distance of $14$). With merge\&shrink
\cite{helmert:etal:icaps07}, we obtained the estimate $11$ for the
initial state, which is much better. Still the search space with that
latter heuristic remained fairly large ($5200$ nodes vs.\ $13800$
nodes for blind search), and it is not clear how well the technique
will scale to larger instances and instances with a more complex
grammar. Certainly, interesting challenges remain.

@@

The only FF technique that appears to be clearly useful here is its
action pruning technique, the ``helpful actions''. While the relaxed
plan is often much too short, it appears to always contain at least
one of the actions that actually are useful. 
%
%As our experiments data
%will show, it certainly prunes the set of available actions quite
%effectively.
%
%For example, in the
%10-distractor test instance, $16875$ nodes are explored without
%helpful actions pruning; with pruning, this reduces to $3078$. For
%larger examples, the factor between the two numbers tends to become
%larger. 
%
%
%Since EHC is often completely inadequate, and helpful actions are
%quite effective, 
%
Given the above observations, we modified FF to directly run
best-first search with helpful actions (as
% regarding EHC and helpful actions, our
%modified FF directly runs best-first search with helpful actions (as
opposed to original FF, which switches to best-first with{\em out}
helpful actions if EHC fails).

### 2009-12-19 trunk/Papers/crisp-ff-10/crisp.tex
4 pages. whew.

\section{Experiences with FF}
\label{sec:crisp-ff}

>>>>

\section{Sentence generation as planning}
\label{sec:crisp}

We consider a planning domain that comes up in the generation of
natural language sentences based on tree-adjoining grammar (TAG;
\cite{joshi;etal1997}).  This is a standard problem in computational
linguistics, which we sketch by example; see
\cite{Stone2003a,KolSto07} for details.

% \begin{figure}
%   \centering
%   \includegraphics[width=0.75\columnwidth]{pic-grammar}
%   \caption{An example grammar in the sentence generation domain.}
%   \label{fig:white-rabbit-sleeps-grammar}
% \end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1\columnwidth]{pic-derivation}
  \caption{Derivation of ``The white rabbit sleeps.''}
  \label{fig:white-rabbit-sleeps-deriv}
\end{figure}

A TAG grammar consists of a finite set of \emph{elementary trees},
each of which contains one or more words; the left of
Fig.~\ref{fig:white-rabbit-sleeps-deriv} shows three trees for ``the
rabbit'', ``sleeps'', and ``white''.  Trees can be combined using the
operations of \emph{substitution} (purple dashed arrow in the figure)
and \emph{adjunction} (blue dotted arrows) to form larger trees.
The end result of a grammatically correct derivation is a tree all of
whose leaves are labeled with words, as on the right of
Fig.~\ref{fig:white-rabbit-sleeps-deriv}. We can read off a sentence
from such a tree from left to right.

To use TAG grammars for generation \cite{Stone2003a}, we assume a set
of ground atoms expressing the information we want the sentence to
convey, such as $\{\mathsf{sleep}(e,r_1)\}$, and a knowledge base that
contains all ground atoms we know to be true; say,
$\{\mathsf{sleep}(e,r_1), \mathsf{rabbit}(r_1), \mathsf{rabbit}(r_2),
\mathsf{white}(r_1), \mathsf{brown}(r_2)\}$.  We then add variables
ranging over constants from the knowledge base to the nodes of each
elementary tree, and equip every elementary tree with a set of atoms
over these variables to encode the meaning this tree expresses.
Fig.~\ref{fig:white-rabbit-sleeps-deriv} already shows specific
instances of the elementary trees from the grammar, in which constants
have been substituted for these variables.  The derivation in the
figure conveys the intended meaning -- in particular, that $r_1$
sleeps.  Crucially, it also describes uniquely who does the sleeping:
The sentence ``the rabbit sleeps'' would not have done this, as ``the
rabbit'' could be understood either as $r_1$ or $r_2$.  We say that
$r_2$ is a \emph{distractor} for the subject of the sentence, which is
removed by adjoining the tree for ``white''.  The sentence generation
problem consists in computing a grammatically correct TAG derivation
from instances of the elementary trees in the grammar that conveys the
intended meaning and describes all referents uniquely.

\subsection{Sentence generation as a planning problem}

The TAG sentence generation problem can be encoded as a planning
problem \cite{KolSto07}.  The key idea is that each operator encodes
the addition of an elementary tree to the derivation; the syntactic
and semantic requirements and effects of doing this are captured in
the operator's precondition and effect.

\newcommand{\action}[4]{\textbf{#1$(#2)$:}\\
\strut\quad   Precond:$\;$ \parbox[t]{12cm}{\ensuremath{#3}}\\
\strut\quad   Effect:$\;$ \parbox[t]{12cm}{\ensuremath{#4}}}
\newcommand{\f}[1]{\mathsf{#1}}

\begin{figure}
%\centering
%\begin{minipage}{0.8\textwidth}
{\small%
\action{sleeps}{u, u_1, u_n, x_0, x_1}{
  \f{subst}(\f{S},u) \wedge \f{ref}(u,x_0) \wedge
  \f{sleep}(x_0,x_1) \\ \wedge \f{current}(u_1) \wedge
  \f{next}(u_1,u_n)
}{
  \neg \f{subst}(\f{S},u) \wedge \f{expressed}(\f{sleep}, x_0, x_1) \\
  \wedge \f{subst}(\f{NP},u) \wedge \f{ref}(u_1,x_1) \\ \wedge
  \neg \f{current}(u_1) \wedge \f{current}(u_n) \\ \wedge
  \forall y. y \neq x_1 \rightarrow \f{distractor}(u_1,y)
}\\

\action{rabbit}{u, x_0}{
  \f{subst}(\f{NP},u) \wedge \f{ref}(u,x_0) \wedge \f{rabbit}(x_0)
}{
  \neg \f{subst}(\f{NP},u) \wedge \f{canadjoin}(\f{N},u) \\
  \wedge \forall y. \neg \f{rabbit}(y) \rightarrow \neg \f{distractor}(u,y)
}\\

\action{white}{u,x_0}{
  \f{canadjoin}(\f{N},u) \wedge \f{ref}(u,x_0) \wedge \f{rabbit}(x_0)
}{
  \forall y. \neg \f{white}(y) \rightarrow \neg \f{distractor}(u,y)
}
}\strut\\[-5ex]
%\end{minipage}
\caption{Actions for generating ``The white rabbit
sleeps.''}
\label{fig:white-rabbit-as-planning}
\end{figure}

More precisely, each operator has a parameter $u$ representing the
syntax node into which the elementary tree is substituted or adjoined,
and a parameter $u_i$ for each substitution node.  There are also
parameters $x_0,\ldots,x_k$ for the variables in the semantic
representation of the elementary tree; $x_0$ is the variable that
occurs at the root of the tree.  The planning state encodes a list of
possible node names; $\f{current}(u)$ expresses that $u$ is the next
node name that should be picked when a new substitution node is
created, and $\f{next}(u,v)$ says that the next node after $u$ is
$v$. These atoms are used to select names for the substitution nodes
that are introduced by adding an elementary tree; the parameter $u_n$
represents the node that will be current after the addition.

The atom $\f{subst}(A,u)$ expresses that we need to still substitute
something into the substitution node $u$ with label $A$; the initial
state contains an atom $\f{subst}(\f{S},\f{root})$ where $\f{S}$
stands for ``sentence'' and $\f{root}$ is an arbitrary name for the
root of the derivation.  The mapping from nodes to semantic constants
is maintained using the $\f{ref}$ atoms; the initial state for
generating a sentence about the event $e$ contains an atom
$\f{ref}(\f{root},e)$.  Finally, we keep track of the uniqueness of
referring expressions in the $\f{distractor}$ atoms:
$\f{distractor}(u,a)$ expresses that the expression at syntax node $u$
could still be misunderstood as $a$.

The example generation problem above translates into a planning
problem whose domain is shown, in PDDL format, in
Fig.~\ref{fig:white-rabbit-as-planning}. The initial state of the
problem encodes the generation problem's knowledge base; it contains
atoms $\f{rabbit}(r_1)$, $\f{sleep}(e,r_1)$, etc.  The goal requires
syntactic completeness as $\forall u \forall A \neg \f{subst}(A,u)$
and unique reference as $\forall u \forall x \neg
\f{distractor}(u,x)$; it also specifies the semantic representation we
want to express in an atom $\f{expressed}(\mathsf{sleep},e,r_1)$.

A minimal plan for the example problem is
$\mathsf{sleeps}(\mathsf{root}, n_1, n_2, e, r_1)$;
$\mathsf{rabbit}(n_1, r_1)$; $\mathsf{white}(n_1, r_1)$. This plan can
be automatically decoded into the derivation shown in
Fig.~\ref{fig:white-rabbit-sleeps-deriv}.  The analogous plan with
$r_2$ instead of $r_1$ is not correct because the $\mathsf{sleep}$
precondition of the $\mathsf{sleeps}$ operator is not satisfied.  The
first two steps of the plan alone would not be a correct plan because
they leave an atom $\mathsf{distractor}(n_1, r_2)$ in the state, which
contradicts the goal that all distractors have been eliminated.

### 2009-12-19 trunk/Papers/crisp-ff-10/experiments.tex
4 pages. whew.

\begin{figure*}[t]
  \centering
\begin{tabular}{cc}
  \includegraphics[width=0.9\columnwidth]{pics/xtag-k2-dist0} & 
\includegraphics[width=0.9\columnwidth]{pics/xtag-k2-dist2}\\[-0.2cm]
(a) & (b)
\end{tabular}
\vspace{-0.3cm}
\caption{Runtimes for $d=0$ (a) and $d=2$ (b).}
  \label{fig:runtimes}
\vspace{-0.5cm}
\end{figure*}

@@

To evaluate the effect of our changes on the efficiency of the planner
on practical inputs, we ran a series of experiments using the XTAG
Grammar \cite{xtag01:_tr}, a large-scale tree-adjoining grammar of
English, from which we selected all lexicon entries for seven words we
used in the experiments. We set up inputs to generate sentences of the
form ``$S_1$ and $S_2$ and \ldots and $S_n$'', i.e.\ a conjunction of
$n$ sentences. Each sentence $S_i$ is of the form ``X admires Y''. X
and Y are unique referring expressions to individuals, such as ``the
rich businessman''. For each experiment, we have a parameter $d$ that
specifies how many adjectives are needed to distinguish the individual
uniquely -- in particular, $d=0$ means we can say ``the businessman'',
and $d=2$ means we must say ``the rich sad businessman''. 
% JOERG: ``new encoding'' gibt's ja gar nicht mehr als unterscheidung
%
%We translate
%these generation problems into planning problems with the new encoding
%as shown in Fig.~\ref{fig:white-rabbit-as-planning}.

@@

We remark that the relative behavior of the different search methods
is not uniform in our domain, which exhibits a huge amount of
structural variance due to the many different forms that the input
grammer may take. We are far from an exhaustive exploration of all the
possible parameter settings. Suffice it to say that, in particular
encoding variants and with particular grammars (e.g.\ grammars not
featuring any distractors), in stark contrast to
Fig.~\ref{fig:runtimes} (b) EHC is much more effective than best-first
search.

>>>>

\begin{figure*}[t]
  \centering
(a)
\vtop{
\vskip-2ex
\hbox{\includegraphics[width=0.8\columnwidth]{pics/xtag-k2-dist0}}
}
$\qquad\qquad$
(b)
\vtop{
\vskip-2ex
\hbox{\includegraphics[width=0.8\columnwidth]{pics/xtag-k2-dist2}}
}
\vspace{-0.5cm}
\caption{Runtimes for $d=0$ (a) and $d=2$ (b).}
  \label{fig:runtimes}
\vspace{-0.5cm}
\end{figure*}

@@

To evaluate the effect of our changes on the efficiency of the planner
on practical inputs, we ran a series of experiments using the XTAG
Grammar \cite{xtag01:_tr}, a large-scale tree-adjoining grammar of
English, from which we selected all lexicon entries for seven words we
used in the experiments. We set up inputs to generate sentences of the
form ``$S_1$ and $S_2$ and \ldots and $S_n$'', i.e.\ a conjunction of
$n$ sentences. Each sentence $S_i$ is of the form ``X admires Y''. X
and Y are unique referring expressions, such as ``the rich
businessman''. For each experiment, we have a parameter $d$ that
specifies how many adjectives are needed to distinguish the referent
uniquely -- in particular, $d=0$ means we can say ``the businessman'',
and $d=2$ means we must say ``the rich sad businessman''.
% JOERG: ``new encoding'' gibt's ja gar nicht mehr als unterscheidung
%
%We translate
%these generation problems into planning problems with the new encoding
%as shown in Fig.~\ref{fig:white-rabbit-as-planning}.

@@

The relative behavior of the different search methods is not uniform
in our domain, which exhibits a huge amount of structural variance due
to the many different forms that the input grammar may take. We are
far from an exhaustive exploration of all the possible parameter
settings. For instance, with particular inputs (no distractors), EHC
can be more effective than best-first search.

%Suffice it to say that, in particular encoding variants and
%with particular grammars (e.g.\ grammars not featuring any
%distractors), in stark contrast to Fig.~\ref{fig:runtimes} (b) EHC is
%much more effective than best-first search.

### 2009-12-19 trunk/Papers/crisp-ff-10/introduction.tex
minor detail
minor
4 pages. whew.

Natural language generation (NLG) -- the problem of computing a
sentence or text that communicates a given piece of information -- has
a long-standing connection to automated planning
\cite{perrault80,appelt:planning}. In the past few years, the idea has
been picked up again in the NLG community, both to use planning as an
approach for modeling language problems
\cite{Steedman-Petrick:07,benotti08b} and to tackle the search
problems in NLG by using techniques from planning \cite{KolSto07}.

@@

In this paper, we revisit this situation, analyze the sources of
inefficiency in more detail in the context of the FF planner
\cite{HoffmannNebel01}, and show how to ``wake up the rabbit'': we
modify FF to obtain dramatically better performance. We achieve this
by tweaking the implementation of FF's preprocessor, and by acting on
some simple but shocking observations regarding the behavior of some
of FF's search techniques, in particular its goal ordering technique
and the relaxed plan heuristic. The changes we make are all simple,
but make FF perform well enough to be useful in practical NLG
applications.

>>>>

Natural language generation (NLG) -- the problem of computing a
sentence or text that communicates a given piece of information -- has
a long-standing connection to automated planning \cite{perrault80}. In
the past few years, the idea has been picked up again in the NLG
community, both to use planning as an approach for modeling language
problems (e.g., \cite{Steedman-Petrick:07}) and to tackle the search
problems in NLG by using techniques from planning \cite{KolSto07}.

@@

In this paper, we revisit this situation, analyze the sources of
inefficiency in more detail in the context of the FF planner
\cite{HoffmannNebel01}, and show how to modify FF to obtain
dramatically better performance. We achieve this by tweaking the
implementation of FF's preprocessor, and by acting on some simple but
shocking observations regarding the behavior of some of FF's search
techniques, in particular its goal ordering technique and the relaxed
plan heuristic. The changes we make are all simple, but make FF
perform well enough to be useful in practical NLG applications.

### 2009-12-19 trunk/Papers/crisp-ff-10/main.tex
experiments v2, added pictures
minor
4 pages. whew.

\begin{document}
\title{Using Planning to Generate Natural-Language Sentences:\\
  Experiences and Challenges}

\author{
Alexander Koller \\ Saarland University, Germany \\ koller@mmci.uni-saarland.de
%\And
%Ronald Petrick \\ University of Edinburgh, UK \\ rpetrick@inf.ed.ac.uk
\And
J\"org Hoffmann \\ INRIA, France \\ joerg.hoffmann@inria.fr
}

>>>>

\begin{document}
\title{Waking up a sleeping rabbit:\\ On Natural-Language Sentence Generation with FF}

%\title{Using Planning to Generate Natural-Language Sentences:\\
%  Experiences and Challenges}

\author{
%Alexander Koller \\ Saarland University, Germany \\ koller@mmci.uni-saarland.de
%\And
%J\"org Hoffmann \\ INRIA, France \\ joerg.hoffmann@inria.fr
}

\nocopyright
